---
title: "Titanic Survivors"
author: "Aisling Mc Mahon 23214497"
date: today
format:
  revealjs:
    center: true
    theme: night
    slide-number: true
    transition: fade
    figure-align: right
    tbl-colwidths: auto
execute: 
  eval: true
  echo: false
  warning: false
  message: false
fontsize: 18pt
---

# The Titanic Dataset

***

###  Titanic Data 

The **Titanic Dataset** is a dataset available from the *carData* package in R. It is a dataframe with 1309 observations and 4 variables. The data contains information on the survival status, sex, age and passenger class of 1309 passengers aboard the Titanic during its fateful voyage in 1912. It contains one numeric variable, age, and three categorical variables, survival, sex and passengerClass.



```{r}
#|label: load_dataset

#load required package

library(carData)
library(dplyr)
library(ggplot2)
library(tidyr)

#load dataset
data("TitanicSurvival")
```


```{r}
#| label: Structure_TitanicSurvival

# look at the structure of the dataset
str(TitanicSurvival)
```

***

### Descriptive statistics for the TitanicSurvival dataset

Before creating a plot, I will look at some descriptive statistics for the **TitanicSurvival** data.



```{r}
#| label: Descriptives_Titanic survival
 
#compile some summary statistics for the dataset

summary(TitanicSurvival)

Percent_sex <- TitanicSurvival|>
  group_by(sex) |> # look at percent male and female
  summarise(Percent = n() / nrow(TitanicSurvival) * 100)

knitr::kable(Percent_sex, digits = 2) 

Percent_survival <- TitanicSurvival|>
  group_by(survived) |> # look at percent who survived and died
  summarise(Percent = n() / nrow(TitanicSurvival) * 100)

knitr::kable(Percent_survival, digits = 2) 

Percent_class <- TitanicSurvival|>
  group_by(passengerClass) |> # look at percent in 1st, 2nd, 3rd class
  summarise(Percent = n() / nrow(TitanicSurvival) * 100)

knitr::kable(Percent_class, digits = 2) 
```


***


To summarise the dataset: 

- the majority of passengers (64.4%) were <span style="color: green;">male</span> 

- the mean age was <span style="color: blue;">29.88</span>, ranging from 2 months old to 80 years old

- just over half (54.16%) being in <span style="color: orange;">3rd class</span> 

- <span style="color: red;">61.8%</span> of passengers on the Titanic died after the ship collided with the ice-berg. 

***

# Plots

***

### Plots

I will now create two plots to visualise survival by ***gender*** and ***class***. The hypothesis would be that survival is higher in females and in first class passengers. This would fit with the premise of women and children first and that passengers in 1st class were likely given priority when lifeboats were being boarded. 

***

### Survival by Gender

:::: {.columns}

::: {.column width="60%"}

```{r}
#| label: Survival_tables
#| fontsize: 12pt

Gender_survival <- TitanicSurvival|>
  group_by(sex)|>
  count(survived) #create a table of survival by sex


Gender_survival <- pivot_wider(Gender_survival,
names_from = survived, values_from = n)

Gender_survival <- Gender_survival |>
mutate(Frequency = yes/(yes+no)) |>
mutate(Percentage = Frequency*100) # calculate proportion and percentage
knitr::kable(Gender_survival, digits = 2) #print table neatly wiith knitr
```
:::

::: {.column width="40%"}

```{r}
#|label: Plots-gender

Mort_gender_plot <- Gender_survival|>
  ggplot(aes(x = sex, y = Percentage, fill = sex)) +
geom_col() + #plot a bar chart for survival by gender
labs(title = "Titanic survival by gender", y = "Percent", x = NULL) +
theme(plot.title = element_text(hjust = 0.5))

Mort_gender_plot #print the plot
```

:::


::::


***

### Survival by Passenger Class


```{r}
#|fontsize: 12pt

Class_survival <- TitanicSurvival|>
  group_by(passengerClass)|>
  count(survived)


Class_survival <- pivot_wider(Class_survival,
names_from = survived, values_from = n)

Class_survival <- Class_survival |>
mutate(Frequency = yes/(yes+no)) |>
mutate(Percentage = Frequency*100)
knitr::kable(Class_survival, digits = 2)

```




```{r}
#|label: Plots-class

Mort_class_plot <- Class_survival|>
  ggplot(aes(x = passengerClass, y = Percentage, fill = passengerClass)) +
geom_col() +
labs(title = "Titanic survival by Class", y = "Percent", x = NULL) +
theme(plot.title = element_text(hjust = 0.5))

Mort_class_plot
```


***

## Conclusion


On inspection on the plots and data from the tables it appears that survival was higher for <span style="color: red;">females</span> passengers and those in <span style="color: green;">1st class</span>.


```{r}
#| label: Stats

chisq.test(TitanicSurvival$sex, TitanicSurvival$survived)
chisq.test(TitanicSurvival$passengerClass, TitanicSurvival$survived)

```

These differences are statistically significant at a p value of <0.05. The null hypothesis that there was no difference in survival based on gender or passenger class can be rejected, and the alternative hypothesis accepted. 

